


## Loading and preprocessing the data

```{r}
library(knitr)
opts_chunk$set(echo = TRUE)
library(ggplot2) 
```
#Load the required data
```{r}
setwd("C:/Users/varshith608/Documents/R/ReproducibleResearch_Project")
rdata <- read.csv("activity.csv")
```
#Tidy the data or preprocess the data
```{r}
rdata$date <- as.Date(rdata$date, format = "%Y-%m-%d")
rdata$interval <- as.factor(rdata$interval)

names(rdata)

str(rdata)
```
## What is mean total number of steps taken per day?

#Lets compute the number of missing vs. complete records

```{r}
originalValue <- complete.cases(rdata)  
nMissing <- length(originalValue[originalValue==FALSE])             # number of records with NA  
nComplete <- length(originalValue[originalValue==TRUE])             # number of complete records
title="Missing vs. Complete Cases"  
barplot(table(originalValue),main=title,xaxt='n', col="gray")       # render Complete Cases barplot  
axis(side=1,at=c(.7,1.9),labels=c("Missing","Complete"),tick=FALSE) # render axis  
text(.7,0,labels=nMissing, pos=3)                                   # label the NA's bar  
text(1.9,0,labels=nComplete, pos=3) 
```

#Now here we ignore the missing values (a valid assumption).
#We proceed by calculating the total steps per day

```{r}
steps_per_day <- aggregate(steps ~ date, rdata, sum)
colnames(steps_per_day) <- c("date","steps")
head(steps_per_day)
```

[1.] Now we make a histogram of the total number of steps taken per day, plotted with appropriate bin interva

```{r}

ggplot(steps_per_day, aes(x = steps)) + 
  geom_histogram(fill = "gray", binwidth = 1000) + 
  labs(title="Histogram of Steps Taken per Day", 
x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 
```

[2.] Now we calculate the mean and median of the number of steps taken per day.

```
steps_mean   <- mean(steps_per_day$steps, na.rm=TRUE)
steps_median <- median(steps_per_day$steps, na.rm=TRUE)
```
The mean is 10766.189 and median is 10765


## What is the average daily activity pattern?
We calculate the aggregation of steps by intervals of 5-minutes and convert the intervals as integers and save them in a data frame called steps_per_interval.

```{r}
steps_per_interval <- aggregate(rdata$steps, 
by = list(interval = rdata$interval),
FUN=mean, na.rm=TRUE)
steps_per_interval$interval <- 
  as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")

```

1. We make the plot with the time series of the average number of steps taken (averaged across all days) versus the 5-minute intervals

```{r}
ggplot(steps_per_interval, aes(x = interval, y = steps)) +   
  geom_line(color="gray", size=1) +  
  labs(title="Average Daily Activity Pattern", x="Interval", y="Number of Steps") +  
  theme_bw()
```

2. Now, we find the 5-minute interval with the containing the maximum number of steps:

```{r}
max_interval <- steps_per_interval[which.max(  
  steps_per_interval$steps),]
```

#The 835th interval has maximum 206 steps.

## Imputing missing values

1. Total number of missing values:
```{r}
missing_vals <- sum(is.na(rdata$steps))
```
2. Strategy for filling in all of the missing values in the dataset
```{r}
na_fill <- function(data, pervalue) {
  na_index <- which(is.na(data$steps))
  na_replace <- unlist(lapply(na_index, FUN=function(idx){
    interval = data[idx,]$interval
    pervalue[pervalue$interval == interval,]$steps
  }))
  fill_steps <- data$steps
  fill_steps[na_index] <- na_replace
  fill_steps
}
```
3. Create a new dataset that is equal to the original dataset but with the missing data filled in
```{r}
rdata_fill <- data.frame(  
  steps = na_fill(rdata, steps_per_interval),  
  date = rdata$date,  
  interval = rdata$interval)
str(rdata_fill)
```
4. A histogram of the total number of steps taken each day
```{r}
fill_steps_per_day <- aggregate(steps ~ date, rdata_fill, sum)
colnames(fill_steps_per_day) <- c("date","steps")
```

##plotting the histogram
```{r}
ggplot(fill_steps_per_day, aes(x = steps)) + 
  geom_histogram(fill = "gray", binwidth = 1000) + 
  labs(title="Histogram of Steps Taken per Day", 
x = "Number of Steps per Day", y = "Number of times in a day (Count)") + theme_bw() 
```

#Calculate and report the mean and median total number of steps taken per day.
```{r}
steps_mean_fill   <- mean(fill_steps_per_day$steps, na.rm=TRUE)
steps_median_fill <- median(fill_steps_per_day$steps, na.rm=TRUE)
```
#Yes, these values do differ slightly.

Before filling the data
  Mean : 10766.189
  Median: 10765
After filling the data
  Mean : 10766.189
  Median: 10766.189
We see that the values after filling the data mean and median are equal.

#What is the impact of imputing missing data on the estimates of the total daily number of steps?
As you can see, comparing with the calculations done in the first section of this document, we observe that while the mean value remains unchanged, the median value has shifted and virtual matches to the mean.

Since our data has shown a t-student distribution (see both histograms), it seems that the impact of imputing missing values has increase our peak, but it’s not affect negatively our predictions.


## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

  1. Augment the table with a column that indicates the day of the week
  2. Subset the table into two parts - weekends (Saturday and Sunday) and weekdays (Monday through Friday).
  3. Tabulate the average steps per interval for each data set.
  4. Plot the two data sets side by side for comparison.
```{r}

weekdays_steps <- function(data) {
  weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                      FUN=mean, na.rm=T)
  # convert to integers for plotting
  weekdays_steps$interval <- 
    as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
  colnames(weekdays_steps) <- c("interval", "steps")
  weekdays_steps
}

data_by_weekdays <- function(data) {
  data$weekday <- 
    as.factor(weekdays(data$date)) # weekdays
  weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
  weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))
  
  weekend_steps <- weekdays_steps(weekend_data)
  weekday_steps <- weekdays_steps(weekday_data)
  
  weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
  weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))
  
  data_by_weekdays <- rbind(weekend_steps, weekday_steps)
  data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
  data_by_weekdays
}

data_weekdays <- data_by_weekdays(rdata_fill)
```
Below you can see the panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends:
  ```{r}
  ggplot(data_weekdays, aes(x=interval, y=steps)) + 
  geom_line(color="darkgray") + 
  facet_wrap(~ dayofweek, nrow=2, ncol=1) +
  labs(x="Interval", y="Number of Steps") +
  theme_bw()
```
It looks like this person may have day job and does most of his or her walking on the weekends!
